---
title: "Report 1D: Expression level,subcellular localization and tissue-specificity of lincRNAs and pcgenes categorized by overlaps with promoters and enhancers."
author: "Cyril Matthey-Doret"
date: "November 18, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#Loading data:
#setwd("/Users/cmatthe5/Documents/First_step/data/")
setwd("/home/cyril/Documents/Master/sem_1/First_step/data/")
#setwd("/home/cyril/Documents/First_step/data/")

# Loading lincRNAs sets
ne_linc <- read.table("enhancer_bound/all_combinations/ne_linc_pr.bed")  # Overlap no enhancer marks, does not take promoters into account
e_linc <- read.table("enhancer_bound/all_combinations/e_linc_pr.bed")  # Overlap enhancer marks, does not take promoters into account
ne.p_linc <- read.table("enhancer_bound/all_combinations/ne.p_linc_pr.bed")  # Overlap promoter marks, but no enhancer marks
e.np_linc <- read.table("enhancer_bound/all_combinations/e.np_linc_pr.bed")  # Overlap enhancer marks, but no promoter marks
e.p_linc <- read.table("enhancer_bound/all_combinations/e.p_linc_pr.bed")  # Overlap promoter marks, and enhancer marks
ne.np_linc <- read.table("enhancer_bound/all_combinations/ne.np_linc_pr.bed")  # Overlap neither promoter marks, nor enhancer marks

# Loading pcgenes sets
ne_pc <- read.table("enhancer_bound/all_combinations/ne_pc_pr.bed")  # Overlap no enhancer marks, does not take promoters into account
e_pc <- read.table("enhancer_bound/all_combinations/e_pc_pr.bed")  # Overlap enhancer marks, does not take promoters into account
ne.p_pc <- read.table("enhancer_bound/all_combinations/ne.p_pc_pr.bed")  # Overlap promoter marks, but no enhancer marks
e.np_pc <- read.table("enhancer_bound/all_combinations/e.np_pc_pr.bed")  # Overlap enhancer marks, but no promoter marks
e.p_pc <- read.table("enhancer_bound/all_combinations/e.p_pc_pr.bed")  # Overlap promoter marks, and enhancer marks
ne.np_pc <- read.table("enhancer_bound/all_combinations/ne.np_pc_pr.bed")  # Overlap neither promoter marks, nor enhancer marks

#adding colnames
colnames(ne_linc)= colnames(e_linc)=colnames(ne.p_linc)=colnames(e.np_linc)=colnames(e.p_linc)=
  colnames(ne.np_linc)=colnames(ne_pc)=colnames(e_pc)=colnames(ne.p_pc)=colnames(e.np_pc)=colnames(e.p_pc)=
  colnames(ne.np_pc) <-c("chr", "start", "end", "gene", "strand")

# Loading genes characteristics
whole_loc <- read.table("../data/subcell_loc/enhancer_promoter/whole_loc.txt", header=T)
whole_exp <- read.table("../data/expression/enhancer_promoter/whole_exp.txt", header=T)
whole_exp <- whole_exp[,c(1,2,6,7,8)]
colnames(whole_exp)[2] <- "expression"
whole_exp <- whole_exp[whole_exp$expression!=0,]
whole_tau <- read.table("../data/tissue_specificity/enhancer_promoter/whole_tau.txt", header=T)
library(ggplot2)
library(gridExtra)
library(plyr)
```


#Introduction

Here, lincRNAs and protein-coding genes were split into categories for each combination of overlap/absence of overlap with promoters and enhancers.

#Data

##lincRNAs

promoter|enhancer|number of genes
------------|-------|-------------
yes|yes|`r length(e.p_linc$chr)`
yes|no|`r length(ne.p_linc$chr)`
no|yes|`r length(e.np_linc$chr)`
no|no|`r length(ne.np_linc$chr)`
both|yes|`r length(e_linc$chr)`
both|no|`r length(ne_linc$chr)`

##protein-coding genes

promoter|enhancer|number of genes
------------|-------|-------------
yes|yes|`r length(e.p_pc$chr)`
yes|no|`r length(ne.p_pc$chr)`
no|yes|`r length(e.np_pc$chr)`
no|no|`r length(ne.np_pc$chr)`
both|yes|`r length(e_pc$chr)`
both|no|`r length(ne_pc$chr)`



#Results

All p-values displayed on plots were obtained by performing Mann-Whitney tests.

##Expression levels




```{r exp_lvl, echo=FALSE,fig.height=8,fig.width=8,results='asis'}

short_med<-function(x){return(round(median(log10(x)),3))}
whole_exp$promoter <- as.factor(whole_exp$promoter)
whole_exp$enhancer <- as.factor(whole_exp$enhancer)
med.fac = ddply(whole_exp, .(gentype, promoter, enhancer), function(.d)
  data.frame(x=median(round(log10(as.numeric(.d$expression)),3))))

ggplot(data=whole_exp[whole_exp$gentype=="linc",])+
  facet_grid(enhancer~promoter)+
  geom_boxplot(aes(x=1, y=log10(expression),fill=enhancer),notch=T)+
  geom_text(data=med.fac[med.fac$gentype=="linc",], aes(x=0.8, y=2, label=paste0("Mdn: ",x)), 
            colour="black", inherit.aes=FALSE, parse=FALSE)+
  theme_bw()+ggtitle("Expression levels of lincRNAs")
cat("\n\n\\pagebreak\n")
ggplot(data=whole_exp[whole_exp$gentype=="pc",])+
  facet_grid(enhancer~promoter)+
  geom_boxplot(aes(x=1, y=log10(expression),fill=enhancer),notch=T)+
  geom_text(data=med.fac[med.fac$gentype=="pc",], aes(x=0.8, y=2, label=paste0("Mdn: ",x)), 
            colour="black", inherit.aes=FALSE, parse=FALSE)+
  theme_bw()+ggtitle("Expression levels of PCGs")
```

##Subcellular localization

The ratios depicted below represent the abundance of RNA in the cytoplasm relative to the nucleus. There is no significant difference between the localization of TAD-bound and non-TAD-bound lincRNAs, while the TAD-bound protein-coding genes are more abundant in the cytoplasm than in the nucleus.


```{r local, echo=FALSE,fig.height=8,fig.width=8,results='asis'}


med.fac = ddply(whole_loc, .(gentype, promoter, enhancer), function(.d)
  data.frame(x=median(round(log10(as.numeric(.d$ratio)),3))))

ggplot(data=whole_loc[whole_loc$gentype=="linc",])+
  facet_grid(enhancer~promoter)+
  geom_boxplot(aes(x=1, y=log10(ratio),fill=enhancer),notch=T)+
  geom_text(data=med.fac[med.fac$gentype=="linc",], aes(x=0.8, y=2, label=paste0("Mdn: ",x)), 
            colour="black", inherit.aes=FALSE, parse=FALSE)+
  ylab("Log10 Cyt/Nuc")+
  theme_bw()+ggtitle("Subcellular localization of lincRNAs")
cat("\n\n\\pagebreak\n")
ggplot(data=whole_loc[whole_loc$gentype=="pc",])+
  facet_grid(enhancer~promoter)+
  geom_boxplot(aes(x=1, y=log10(ratio),fill=enhancer),notch=T)+
  geom_text(data=med.fac[med.fac$gentype=="pc",], aes(x=0.8, y=2, label=paste0("Mdn: ",x)), 
            colour="black", inherit.aes=FALSE, parse=FALSE)+
  ylab("Log10 Cyt/Nuc")+
  theme_bw()+ggtitle("Subcellular localization of PCGs")
```

##Tissue specificity

$\tau$ was calculated from a set of tissue-wise expression level for all transcripts. Again, there seem to be no difference in tissue specificity between TAD-bound and nonTAD-bound lincRNAs or protein coding genes. Overall the lincRNAs are more tissue specific than protein coding genes.

```{r tissue, echo=FALSE,fig.height=8,fig.width=8,results='asis'}


med.fac = ddply(whole_tau, .(gentype, promoter, enhancer), function(.d)
  data.frame(x=median(round(as.numeric(.d$tau),3))))

ggplot(data=whole_tau[whole_tau$gentype=="linc",])+
  facet_grid(enhancer~promoter)+
  geom_boxplot(aes(x=1, y=tau,fill=enhancer),notch=T)+
  geom_text(data=med.fac[med.fac$gentype=="linc",], aes(x=0.8, y=1.1, label=paste0("Mdn: ",x)), 
            colour="black", inherit.aes=FALSE, parse=FALSE)+
  ylab("Tau")+
  theme_bw()+ggtitle("Tissue specificity of lincRNAs")
cat("\n\n\\pagebreak\n")
ggplot(data=whole_tau[whole_tau$gentype=="pc",])+
  facet_grid(enhancer~promoter)+
  geom_boxplot(aes(x=1, y=tau,fill=enhancer),notch=T)+
  geom_text(data=med.fac[med.fac$gentype=="pc",], aes(x=0.8, y=1.1, label=paste0("Mdn: ",x)), 
            colour="black", inherit.aes=FALSE, parse=FALSE)+
  ylab("Tau")+
  theme_bw()+ggtitle("Tissue specificity of PCGs")

```


