---
title: "Report 1: Expression level,subcellular localization and tissue-specificity of TADbound and non-TADbound lincRNAs."
author: "Cyril Matthey-Doret"
date: "October 6, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#Loading data:
setwd("/home/cyril/Documents/Master/sem_1/First_step/data/")
#setwd("/home/cyril/Documents/First_step/data/")
whole_loc <- read.table("subcell_loc/whole_loc.txt", header=T)
whole_exp <- read.table("expression/whole_exp.txt", header=T)
whole_tau <- read.table("tissue_specificity/whole_tau.txt", header=T)
library(ggplot2)
library(gridExtra)
library(plyr)
```
#TAD boundaries definition
When defining TAD boundaries as the areas 20% before and after TAD borders, those boundaries seemed to overlap with those of the neighbouring domains. To solve this issue, the TAD boundaries have been set to 3 different, arbitrary threshold: 5%, 10% and 20% of the TAD length.

#Overlapping genes with TAD boundaries
Both lincRNAs and protein coding genes have been overlapped with TAD boundaries, using the intersect program with -f 0.25 (genes must have at least 25% of their sequence overlapping a boundary to be considered). The set of overlaps was then cleaned by removing duplicates.

#Data

##lincRNA

Original set|Threshold|Overlaps|TAD-bound genes
------------|---------|--------|-----
2510|5%|865|609 (`r round((100*609)/2510,1)`%)
||10%|1481|960 (`r round((100*960)/2510,1)`%)
||20%|2554|1420 (`r round((100*1420)/2510,1)`%)

##Protein-coding genes

Original set|Threshold|Overlaps|TAD-bound genes
------------|---------|--------|-----
14846|5%|6576|4607 (`r round((100*4607)/14846,1)`%)
||10%|12017|7495 (`r round((100*7495)/14846,1)`%)
||20%|21544|11337 (`r round((100*11337)/14846,1)`%)

#Results

All p-values displayed on plots were obtained by performing Mann-Whitney tests.

##Expression levels

It seems the threshold used to define TAD-boundariesdoes not have much influence on the results.
The TAD-bound lincRNAs have similar expression levels to the non-TAD-bound lincRNAs, regardless of which threshold was used, whereas the TAD-bound protein-coding genes where always more expressed than the rest.


```{r exp_lvl, echo=FALSE,fig.height=8,fig.width=8}
short_med<-function(x){return(round(median(log10(x)),3))}
short_wilcox <- function(x,y){return(format(wilcox.test(x, y)$p.value,digits=3))}
whole_exp$threshold <- as.factor(whole_exp$threshold)
med.fac = ddply(whole_exp, .(gentype, threshold,TAD), function(.d)
  data.frame(x=median(round(log10(.d$expression),3))))
pl<-c()
gl<-c()
tl<-c()
for(t in levels(as.factor(whole_exp$threshold))){
  for(g in levels(whole_exp$gentype)){
    p <- short_wilcox(whole_exp[whole_exp$threshold==t & whole_exp$gentype==g &  whole_exp$TAD=="Tb","expression"],
                      whole_exp[whole_exp$threshold==t &  whole_exp$gentype==g & whole_exp$TAD=="nTb","expression"])
    pl<-append(pl,p)
    tl<-append(tl,t)
    gl<-append(gl,g)
  }
}
wilcox_p <- data.frame(pval=pl,threshold=tl,gentype=gl)

ggplot(data=whole_exp)+
  facet_grid(gentype~threshold)+
  geom_boxplot(aes(x=TAD, y=log10(expression),fill=TAD),notch=T)+
  geom_text(data=med.fac, aes(x=TAD, y=-1, label=x), 
            colour="black", inherit.aes=FALSE, parse=FALSE)+
  geom_text(data=wilcox_p, aes(x=1.5, y=1, label=paste0("p=",pval)), 
            colour="black", inherit.aes=FALSE, parse=FALSE)+
  theme_bw()
```

##Subcellular localization

The ratios depicted below represent the abundance of RNA in the cytoplasm relative to the nucleus. Again, regardless of the threshold used to define TAD-boundaries, There is no significant difference between the localization of TAD-bound and non-TAD-bound lincRNAs, while the TAD-bound protein-coding genes are always more abundant in the cytoplasm than in the nucleus.


```{r local, echo=FALSE,fig.height=8,fig.width=8}

short_med<-function(x){return(round(median(log10(x)),3))}
short_wilcox <- function(x,y){return(format(wilcox.test(x, y)$p.value,digits=3))}
whole_loc$threshold <- as.factor(whole_loc$threshold)

med.fac = ddply(whole_loc, .(gentype, threshold,TAD), function(.d)
  data.frame(x=round(median(log10(.d$ratio)),3)))
#data frame for annotation of p-value
pl<-c()
gl<-c()
tl<-c()
for(t in levels(whole_loc$threshold)){
  for(g in levels(whole_loc$gentype)){
    p <- short_wilcox(whole_loc[whole_loc$threshold==t & whole_loc$gentype==g &  whole_loc$TAD=="Tb","ratio"],
                      whole_loc[whole_loc$threshold==t &  whole_loc$gentype==g & whole_loc$TAD=="nTb","ratio"])
    pl<-append(pl,p)
    tl<-append(tl,t)
    gl<-append(gl,g)
  }
}
wilcox_p <- data.frame(pval=pl,threshold=tl,gentype=gl)
library(ggplot2)
ggplot(data=whole_loc)+
  facet_grid(gentype~threshold)+
  geom_boxplot(aes(x=TAD, y=log10(ratio),fill=TAD),notch=T)+
  geom_text(data=med.fac, aes(x=TAD, y=-1, label=x), 
            colour="black", inherit.aes=FALSE, parse=FALSE)+
  geom_text(data=wilcox_p, aes(x=1.5, y=0.5, label=paste0("p=",pval)), 
            colour="black", inherit.aes=FALSE, parse=FALSE)+
  theme_bw()

```

##Tissue specificity

$\tau$ was calculated from a set of tissue-wise expression level for all transcripts. Again, there seem to be no difference in tissue specificity between TAD-bound and nonTAD-bound lincRNAs, whereas the TAD-bound protein coding genes appear to be less tissue specific. Overall the lincRNAs are more tissue specific than protein coding genes.

```{r tissue, echo=FALSE,fig.height=8,fig.width=8}

library(plyr)
short_med<-function(x){return(round(median(log10(x)),3))}
short_wilcox <- function(x,y){return(format(wilcox.test(x, y)$p.value,digits=3))}
whole_tau$threshold <- as.factor(whole_tau$threshold)
#dataf frame for annotation of median
med.fac = ddply(whole_tau, .(gentype, threshold,TAD), function(.d)
  data.frame(x=median(round(.d$tau,3))))
#data frame for annotation of p-value
pl<-c()
gl<-c()
tl<-c()
for(t in levels(whole_tau$threshold)){
  for(g in levels(whole_tau$gentype)){
    p <- short_wilcox(whole_tau[whole_tau$threshold==t & whole_tau$gentype==g &  whole_tau$TAD=="Tb","tau"],
                      whole_tau[whole_tau$threshold==t &  whole_tau$gentype==g & whole_tau$TAD=="nTb","tau"])
    pl<-append(pl,p)
    tl<-append(tl,t)
    gl<-append(gl,g)
  }
}
wilcox_p <- data.frame(pval=pl,threshold=tl,gentype=gl)
library(ggplot2)
ggplot(data=whole_tau)+
  facet_grid(gentype~threshold)+
  geom_boxplot(aes(x=TAD, y=tau,fill=TAD),notch=T)+
  geom_text(data=med.fac, aes(x=TAD, y=0, label=x), 
            colour="black", inherit.aes=FALSE, parse=FALSE)+
  geom_text(data=wilcox_p, aes(x=1.5, y=0.25, label=paste0("p=",pval)), 
            colour="black", inherit.aes=FALSE, parse=FALSE)+
  theme_bw()
```

#Conclusion

The TAD definition might not be appropriate. It was already pointed out (Rao et al, 2014) that the large TADs often contain mainy smaller TADs (see pictures from UCSC). These overlaps
often cause the TAD boundaries to overlap and probably generate a lot of noise in the results. It might be more interesting to merge overlapping TADs in order to consider only the largest ones and their boundaries.