---
title: 'Report 5: Seqence conservation'
author: "Cyril Matthey-Doret"
date: "25 octobre 2016"
output: pdf_document
linkcolor: "blue"
---

```{r setup, include=FALSE}
#setwd("/home/cyril/Documents/Master/sem_1/First_step/data/")
setwd("/home/cyril/Documents/First_step/data/")
#setwd("/Users/cmatthe5/Documents/First_step/data/")
whole_cons <- read.table("../data/seq_conserv/enhancer_bound/whole_cons.txt", header=T)
```

#Introduction

Sequence conservation across species can be quantified using [__phastCons scores__](http://genome.cshlp.org/content/15/8/1034.full). Here I compare averaged sequence conservation of enhancer-associated (e) and promoter-associated (p) lincRNAs and protein-coding genes. Ancestral repeats (AR) are also used as a comparison, since those elements are assumed to be evolving neutrally. Genes that have a higher phastCons score than ancestral repeats are likely to be under purifying selection, while those with a lower score should be under positive selection.


#Comparison of sequence conservation

```{r cons_comp, echo=F, error=F,warning=F,fig.width=8,fig.height=6}
library(ggplot2);library(gridExtra);library(plyr);library(scales)
options(digits=3,scipen=0)
short_med<-function(x){return(round(median(x),3))}
short_wilcox <- function(x,y){return(format(wilcox.test(x, y)$p.value,digits=3))}
#dataf frame for annotation of median
med.fac <- ddply(whole_cons, .(gentype, gr,assoc), function(.d)
  data.frame(x=median(na.rm = T,round(.d$avg_score,3))))
#data frame for annotation of p-value
pl<-c()
gl<-c()
tl<-c()
cons <- whole_cons[whole_cons$gentype!="AR",]
cons$gentype <- droplevels(cons$gentype);cons$assoc <- droplevels(cons$assoc)
for(t in levels(cons$gr)){
  for(g in levels(cons$gentype)){
    p <- short_wilcox(cons[cons$gr==t & cons$gentype==g &  cons$assoc=="e","avg_score"],
                      cons[cons$gr==t &  cons$gentype==g & cons$assoc=="p","avg_score"])
    p <- as.numeric(p)
    if(nchar(p)>4){p <-format(p,scientific=T)}
    pl<-append(pl,p)
    tl<-append(tl,t)
    gl<-append(gl,g)
  }
}
wilcox_p <- data.frame(pval=pl,gr=tl,gentype=gl)

dl <- ggplot(data = whole_cons,aes(x=avg_score),stat=..density..)+
  geom_histogram(data = whole_cons[whole_cons$gentype=="lincRNA" & whole_cons$assoc=="e",],bins=60,fill="#bb6666")+
  geom_histogram(data = whole_cons[whole_cons$gentype=="lincRNA" & whole_cons$assoc=="p",],bins=60,fill="#6666bb")+
  theme_bw()+ggtitle("lincRNAs")+theme(axis.title=element_blank())

dp <- ggplot(data = whole_cons,aes(x=avg_score),stat=..density..)+
  geom_histogram(data = whole_cons[whole_cons$gentype=="pc" & whole_cons$assoc=="e",],bins=60,fill="#bb6666")+
  geom_histogram(data = whole_cons[whole_cons$gentype=="pc" & whole_cons$assoc=="p",],bins=60,fill="#6666bb")+
  theme_bw()+ggtitle("Protein-coding genes")+theme(axis.title=element_blank())

dr <- ggplot(data = whole_cons[whole_cons$gentype=="AR",],aes(x=avg_score))+
  geom_histogram(bins=60,fill="#66bb66")+ggtitle("Ancestral repeats")+
  theme_bw()+theme(axis.title.y=element_blank())+xlab("averaged phastCons")

group_names <- c(`mam` = "mammals",`pri` = "primates")
gentype_names <- c(`lincRNA` = "lincRNAs",`pc` = "protein-coding")
AR_medcons <- data.frame(val=c(median(whole_cons$avg_score[whole_cons$gentype=="AR" & whole_cons$gr=="mam"],na.rm = T),
                               median(whole_cons$avg_score[whole_cons$gentype=="AR" & whole_cons$gr=="mam"],na.rm = T),
                               median(whole_cons$avg_score[whole_cons$gentype=="AR" & whole_cons$gr=="pri"],na.rm = T),
                               median(whole_cons$avg_score[whole_cons$gentype=="AR" & whole_cons$gr=="pri"],na.rm = T)),
                         gentype=c("lincRNA","pc","lincRNA","pc"),gr=c("mam","mam","pri","pri"))

c <- ggplot(data=whole_cons[whole_cons$gentype!="AR",])+
  facet_grid(gr~gentype,labeller = labeller(gentype = as_labeller(gentype_names),
                                                 gr = as_labeller(group_names)))+
  geom_boxplot(aes(x=assoc,y=avg_score,fill=assoc),notch=T)+
  geom_hline(data= AR_medcons,aes(yintercept = AR_medcons$val,fill="AR"),size=1,col="#66bb66",show.legend= T)+
  scale_fill_manual(values = c("#66bb66","#bb6666","#6666bb"))+
  scale_linetype_manual("Title", values = 2) +
  guides(fill = guide_legend(title = element_blank(),override.aes=list(linetype=0)))+
  geom_text(data=wilcox_p, aes(x=1.7, y=1, label=paste0("p=",pval)), 
            colour="black", inherit.aes=FALSE, parse=FALSE)+
  geom_label(data=med.fac[med.fac$gentype!="AR",], aes(x=assoc, y=x+0.01, label=x), 
            colour="black", inherit.aes=FALSE, parse=FALSE,size=3)+
  theme_bw()+ ylab("averaged phastCons")

grid.arrange(layout_matrix=matrix(c(1,4,4,2,4,4,3,4,4),nrow = 3,byrow = T),grobs = list(dl,dp,dr,c))



```
