---
title: 'Report 3: Redefining boundaries'
author: "Cyril Matthey-Doret"
date: "11 octobre 2016"
output: pdf_document
---

```{r setup, include=FALSE}
#Loading data
setwd("/home/cyril/Documents/Master/sem_1/First_step/data/")
#setwd("/home/cyril/Documents/First_step/data/")
#setwd("/Users/cmatthe5/Documents/First_step/data/")
library(ggplot2)
library(gridExtra)
m.TAD<- read.table("../data/TAD/merged/merged_TAD.bed") #merged TADs, or "Large TADs"
m.TAD <- cbind(paste("merged",row.names(m.TAD),sep="_"),m.TAD) #adding arbitrary ID to TADs just in case
TAD <- read.table("../data/TAD/GM12878_TAD_domains.bed") #original TADs, high number of overlapping domains
TAD <- TAD[,-4]
colnames(TAD) <- c("chr","start","end","ID")
m.gap <- read.table("../data/TAD/merged/gaps.bed") #gaps between merged TADs
colnames(m.TAD) = colnames(m.gap) <- c("ID","chr","start","end")
m.TAD_b5 <- read.table("../data/TAD/merged/TAD_boundaries5.bed")
m.TAD_b10 <- read.table("../data/TAD/merged/TAD_boundaries10.bed")
m.TAD_b20 <- read.table("../data/TAD/merged/TAD_boundaries20.bed")
m.TADbins <- read.table("../data/TAD/merged/merged_TADbins.txt")
m.TADbound <- read.table("../data/TAD/merged/whole_TAD_boundaries.txt")
colnames(m.TADbound)=colnames(m.TADbins) <- c("ID","chr","pos","start","end")
RNAcount_bin10 <- read.table("../data/TAD/merged/RNAcount_TADbins10.txt")
RNAcount_bin5 <- read.table("../data/TAD/merged/RNAcount_TADbins5.txt")
RNAcount_bin1 <- read.table("../data/TAD/merged/RNAcount_TADbins1.txt")
colnames(RNAcount_bin)=colnames(RNAcount_bin5)=colnames(RNAcount_bin1)<-
  c("ID","chr","pos","start","end","lincRNA","pcgene")
  
```


#Introduction

Visualising and quantifying the distribution of TAD lengths, or the areas of ovelap between lincRNAs and TADs might help finding a better threshold for the definition of TAD boundaries. This report aims to provide support for the definition of those boundaries.

#1. Length distribution of TADs, boundaries and gaps
   + Length of TADs:
   TADs often contained many smaller TADs which causes the boundaries to overlap and makes it harder to distinguish between TAD-bound and non-TAD-bound RNAs. All overlapping TADs were merge so that only large domains would be considered and there would not be any boundaries contained inside another domain. The mean of the length distribution for the newly generated TADs (Large TADs) is significantly higher.
   
```{r TAD_length, echo=F, warning=F, error=F, fig.height=5, fig.width=6}

options(scipen=4)
dist_TAD <-ggplot(data=TAD)+
  geom_histogram(aes(x=log10(end-start),y=..count..),fill="#5555cc",bins = 30)+
  coord_cartesian(xlim = c(4.5,6.5))+
  xlab(label = "")+
  theme_bw()
dist_mTAD <-ggplot(data=m.TAD)+
  geom_histogram(aes(x=log10(end-start),y=..count..),fill="#55cc55",bins = 30)+
  coord_cartesian(xlim = c(4.5,6.5))+
  xlab(label = "Log10 Length")+
  theme_bw()
dist_TADvsmTAD <-ggplot()+
  geom_boxplot(data=TAD,aes(x="original TADs",y=log10(end-start)),fill="#5555cc",notch = T)+
  geom_boxplot(data=m.TAD,aes(x = "Large TADs",y=log10(end-start)),fill="#55cc55",notch = T)+
  ylab(label = "Log10 Length")+
  annotate(geom = "label",x = c(1,2),y=6,label=c(paste0("Mdn=",format(as.numeric(median(m.TAD$end-m.TAD$start)),scientific=T)),
                                                 paste0("Mdn=",format(median(TAD$end-TAD$start),scientific = T))))+
  xlab(paste0("p-value=",format(wilcox.test(m.TAD$end-m.TAD$start,TAD$end-TAD$start)$p.value,digits=2)))+
  theme_bw()
grid.arrange(layout_matrix = matrix(c(1,3,2,3),nrow = 2,byrow = T),grobs = list(dist_TAD,dist_mTAD,dist_TADvsmTAD))
options(scipen=0)

```

   + Length of TAD boundaries:
   TAD boundaries have been redefined so that they are more flexible. Again, three different thresholds have been used to define them, however a boundary cannot overlap another. A TAD boundary will always reach only as far as the smallest value between the threshold and half the gap between its TAD and the neighbour.
   
```{r boundaries_length, echo=F, warning=F, error=F, fig.height=4, fig.width=5}
# Splitting position into 2 factors so that they can be used to facet the plots.
library(stringi)
get.side <-function(TADbound){
  side <- substr(TADbound[3],1,1)
  return(side)
}
get.threshold <-function(TADbound){
  threshold <- stri_sub(TADbound[3],2,-1)
  return(threshold)
}
side <- apply(m.TADbound, MARGIN = 1,FUN = get.side)
thresh <- apply(m.TADbound, MARGIN = 1,FUN = get.threshold)
ext.m.TADbound <- data.frame(m.TADbound,side=side,threshold=thresh)
med.TADbound <- by(ext.m.TADbound$end-ext.m.TADbound$start,INDICES = relevel(ext.m.TADbound$threshold,"5"),median)
ggplot(data=ext.m.TADbound)+
  geom_boxplot(aes(x=relevel(threshold,"5"),y=log10(end-start)),fill=c("#aa4444","#44aa44","#4444aa"))+
  annotate(geom = "label",x=c(1,2,3),y=log10(c(med.TADbound[1:3])),label=c(paste0("Mdn=\n",med.TADbound[1:3])))+
  xlab("Threshold")+
  ylab("Boundary length")+
  ggtitle("Length of redefined boundaries by threshold")+
  theme_bw()

```

   + Length of gaps: 
   The distribution gap length show that the gaps are generally smaller than the TADs. The median length of gaps is `r round(median(m.TAD$end-m.TAD$start)/median(m.gap$end-m.gap$start),2)` times smaller than for TADs. The discrete distribution observed on a log scale is caused by the approximations that are used for defining the border of TADs.
   
```{r gap_length, echo=F, warning=F,error=F,fig.with=5,fig.height=3}

ggplot(data=m.gap,aes(x = log10(end-start)))+
  geom_histogram(bins = 50, fill="steelblue")+
  ggtitle("Distribution of gap length")+
  xlab("Log10 length")+
  geom_vline(xintercept = log10(median(m.gap$end - m.gap$start)),col="red")+
  annotate(geom="label",label=paste0("Median length: ",median(m.gap$end - m.gap$start)),x=log10(median(m.gap$end - m.gap$start))+0.6,y=300)+
  theme_bw()

```

#2. Overlaps between RNA and areas of TADs

TADs were divided into 10 bins, plus 3 outer bins on both sides. The length of each inner bin is 10% of its TAD length. For outer bins, the length was adjusted so that it would not overlap with the next TAD. If half the distance until the next TAD was larger than 30% (i.e. 3 bins of 10%) of TAD length, the size of outer bins remains unchanged. Otherwise, half the gap between neighbouring TADs is split into the 3 bins, meaning each bin is resized to one sixth of its TAD length.

In order to know in which area of TADs lincRNAs and protein coding genes are more likely to be located, all of these genes with at least 25% of their sequence overlapping a bin were counted. The total number of lincRNAs or protein-coding genes that overlap with each bin was then calculated. 

Using the raw number of genes overlapping each bin would be biased since longer TADs have have longer bins, and those bins will have more weight on the plot. The barplots below show the values after they have been normalized by bin length. The bins depicted in orange are located outside TADs, while the ones in blue are inside. All bins beginning by L are on the left (before the TAD), while those beginning with R are on the right (after the TAD).



```{r normalized_overlap_bins, echo=F,warning=F,error=F,fig.width=6,fig.height=5,}

short_RNAcount_bin10 <- data.frame(pos=rep(0,length(levels(RNAcount_bin10$pos))),
                                 pcgene=rep(0,length(levels(RNAcount_bin10$pos))),
                                 lincRNA=rep(0,length(levels(RNAcount_bin10$pos))))
c <- 1
for(i in levels(RNAcount_bin10$pos)){
  pc_short_count <- sum(RNAcount_bin10[RNAcount_bin10$pos==i,"pcgene"]/(RNAcount_bin10[RNAcount_bin10$pos==i,"end"]
                                                                     -RNAcount_bin10[RNAcount_bin10$pos==i,"start"]))
  lincRNA_short_count <- sum(RNAcount_bin10[RNAcount_bin10$pos==i,"lincRNA"]/(RNAcount_bin10[RNAcount_bin10$pos==i,"end"]
                                                                     -RNAcount_bin10[RNAcount_bin10$pos==i,"start"]))
  short_RNAcount_bin10[c,] <- cbind(i,pc_short_count,lincRNA_short_count)
  c <- c+1
}

short_RNAcount_bin10$pos <- factor(short_RNAcount_bin10$pos,c("L3","L2","L1",seq(from=10,to=100,by=10),"R1","R2","R3"),ordered=T)

l <- ggplot(data = short_RNAcount_bin10, aes(x=pos,y=as.numeric(lincRNA),fill=pos))+
  geom_bar(stat="identity")+
  scale_fill_manual(values=c(rep("#E69F00",3),rep("#6666E6",10),rep("#E69F00",3)))+
  guides(fill=F)+
  xlab("TAD bin")+
  ylab("Normalized N lincRNAs")+
  ggtitle("lincRNA per base per TAD bin")+
  theme_bw()
p <- ggplot(data = short_RNAcount_bin10, aes(x=pos,y=as.numeric(pcgene),fill=pos))+
  geom_bar(stat="identity")+
  scale_fill_manual(values=c(rep("#E69F00",3),rep("#6666E6",10),rep("#E69F00",3)))+
  guides(fill=F)+
  xlab("TAD bin")+
  ylab("Normalized N p.c. genes")+
  ggtitle("Protein-coding gene per base per TAD bin")+
  theme_bw()
grid.arrange(l,p)

# Looks cool but doesn't work properly:
#gg <- ggplot(RNAcount_bin10, aes(x=chr, y=pos, fill=pcgene))
#gg <- gg + geom_tile(color="white", size=0.1)
#gg <- gg + scale_fill_viridis(name="# pc genes")
#gg <- gg + coord_equal()

```

Below are the same barplots, using respectively 5% and 1% for binwidth instead of 10%. Having different resolutions for the detection of genes location in TAD boundaries will help defining a better threshold for the TAD boundaries.


```{r overlap_bins5, echo=F,warning=F,error=F,fig.width=6,fig.height=5,}

short_RNAcount_bin5 <- data.frame(pos=rep(0,length(levels(RNAcount_bin5$pos))),
                                 pcgene=rep(0,length(levels(RNAcount_bin5$pos))),
                                 lincRNA=rep(0,length(levels(RNAcount_bin5$pos))))
c <- 1
for(i in levels(RNAcount_bin5$pos)){
  pc_short_count <- sum(RNAcount_bin5[RNAcount_bin5$pos==i,"pcgene"]/(RNAcount_bin5[RNAcount_bin5$pos==i,"end"]
                                                                     -RNAcount_bin5[RNAcount_bin5$pos==i,"start"]))
  lincRNA_short_count <- sum(RNAcount_bin5[RNAcount_bin5$pos==i,"lincRNA"]/(RNAcount_bin5[RNAcount_bin5$pos==i,"end"]
                                                                     -RNAcount_bin5[RNAcount_bin5$pos==i,"start"]))
  short_RNAcount_bin5[c,] <- cbind(i,pc_short_count,lincRNA_short_count)
  c <- c+1
}

short_RNAcount_bin5$pos <- factor(short_RNAcount_bin5$pos,c(paste0("L",seq(from=6,to=1)),
                                                            seq(from=10,to=100,by=10),
                                                            paste0("R",seq(from=1,to=6))),ordered=T)

l <- ggplot(data = short_RNAcount_bin5, aes(x=pos,y=as.numeric(lincRNA),fill=pos))+
  geom_bar(stat="identity")+
  scale_fill_manual(values=c(rep("#E69F00",3),rep("#6666E6",10),rep("#E69F00",3)))+
  guides(fill=F)+
  xlab("TAD bin")+
  ylab("Normalized N lincRNAs")+
  ggtitle("lincRNA per base per TAD bin")+
  theme_bw()
p <- ggplot(data = short_RNAcount_bin5, aes(x=pos,y=as.numeric(pcgene),fill=pos))+
  geom_bar(stat="identity")+
  scale_fill_manual(values=c(rep("#E69F00",3),rep("#6666E6",10),rep("#E69F00",3)))+
  guides(fill=F)+
  xlab("TAD bin")+
  ylab("Normalized N p.c. genes")+
  ggtitle("Protein-coding gene per base per TAD bin")+
  theme_bw()
grid.arrange(l,p)


```

```{r overlap_bins1, echo=F,warning=F,error=F,fig.width=6,fig.height=5,}

short_RNAcount_bin1 <- data.frame(pos=rep(0,length(levels(RNAcount_bin1$pos))),
                                 pcgene=rep(0,length(levels(RNAcount_bin1$pos))),
                                 lincRNA=rep(0,length(levels(RNAcount_bin1$pos))))
c <- 1
for(i in levels(RNAcount_bin1$pos)){
  pc_short_count <- sum(RNAcount_bin1[RNAcount_bin1$pos==i,"pcgene"]/(RNAcount_bin1[RNAcount_bin1$pos==i,"end"]
                                                                     -RNAcount_bin1[RNAcount_bin1$pos==i,"start"]))
  lincRNA_short_count <- sum(RNAcount_bin1[RNAcount_bin1$pos==i,"lincRNA"]/(RNAcount_bin1[RNAcount_bin1$pos==i,"end"]
                                                                     -RNAcount_bin1[RNAcount_bin1$pos==i,"start"]))
  short_RNAcount_bin1[c,] <- cbind(i,pc_short_count,lincRNA_short_count)
  c <- c+1
}

short_RNAcount_bin1$pos <- factor(short_RNAcount_bin1$pos,c(paste0("L",seq(from=6,to=1)),
                                                            seq(from=10,to=100,by=10),
                                                            paste0("R",seq(from=1,to=6))),ordered=T)

l <- ggplot(data = short_RNAcount_bin1, aes(x=pos,y=as.numeric(lincRNA),fill=pos))+
  geom_bar(stat="identity")+
  scale_fill_manual(values=c(rep("#E69F00",3),rep("#6666E6",10),rep("#E69F00",3)))+
  guides(fill=F)+
  xlab("TAD bin")+
  ylab("Normalized N lincRNAs")+
  ggtitle("lincRNA per base per TAD bin")+
  theme_bw()
p <- ggplot(data = short_RNAcount_bin1, aes(x=pos,y=as.numeric(pcgene),fill=pos))+
  geom_bar(stat="identity")+
  scale_fill_manual(values=c(rep("#E69F00",3),rep("#6666E6",10),rep("#E69F00",3)))+
  guides(fill=F)+
  xlab("TAD bin")+
  ylab("Normalized N p.c. genes")+
  ggtitle("Protein-coding gene per base per TAD bin")+
  theme_bw()
grid.arrange(l,p)


```


#3. ...