---
title: "Report: Expression level and subcellular localization of TADbound and non-TADbound lincRNAs."
author: "Cyril Matthey-Doret"
date: "October 6, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/home/cyril/Documents/First_step/data/")
library(ggplot2)
```
#TAD boundaries definition
When defining TAD boundaries as the areas 20% before and after TAD borders, those boundaries seemed to overlap with those of the neighbouring domains. To solve this issue, the TAD boundaries have been set to 3 different, arbitrary threshold: 5%, 10% and 20% of the TAD length.

#Overlapping genes with TAD boundaries
Both lincRNAs and protein coding genes have been overlapped with TAD boundaries, using the intersect program with -f 0.25 (genes must have at least 25% of their sequence overlapping a boundary to be considered). The set of overlaps was then cleaned by removing duplicates.

#Data

##lincRNA

Original set|Threshold|Overlaps|TAD-bound genes
------------|---------|--------|-----
2510|5%|865|609
||10%|1481|960
||20%|2554|1420

#Protein-coding genes

Original set|Threshold|Overlaps|TAD-bound genes
------------|---------|--------|-----
14846|5%|6576|4607
||10%|12017|7495
||20%|21544|11337

#Results

##Expression levels

###lincRNA

```{r exp_lincRNA, echo=FALSE}
exp_Tb_lincRNA <- read.table("data/expression/")
boxplot(log10(exp_nTb_lincRNA$expression), log10(exp_Tb_pc$expression),notch = T)
ggplot()+
  geom_boxplot()
```

###Protein-coding genes

##Subcellular localization

...

##Tissue specificity

...




