---
title: "Report: Expression level and subcellular localization of TADbound and non-TADbound lincRNAs."
author: "Cyril Matthey-Doret"
date: "October 6, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/home/cyril/Documents/First_step/data/")
#Loading data:
setwd("/home/cyril/Documents/First_step/data/")
#setwd("/home/cyril/Documents/Master/sem_1/First step/data/")
bedcol <- c("chr", "start", "end", "gene", "strand")
exp_lincRNA <- read.table("expression/LCL.lincRNA.expression.txt", header = F)
exp_pcgene <- read.table("expression/LCL.pcgene.expression.txt", header = F)
#loading TAD-bound lincRNAs sets
Tb_lincRNA5 <- read.table("linc_RNA/TADbound-lincRNA5.bed")
Tb_lincRNA10 <- read.table("linc_RNA/TADbound-lincRNA10.bed")
Tb_lincRNA20 <- read.table("linc_RNA/TADbound-lincRNA20.bed")
#loading TAD-bound pcgenes sets
Tb_pc5 <- read.table("pc_genes/TADbound-pcgene5.bed")
Tb_pc10 <- read.table("pc_genes/TADbound-pcgene10.bed")
Tb_pc20 <- read.table("pc_genes/TADbound-pcgene20.bed")
#loading non-TAD-bound lincRNAs sets
nTb_lincRNA5 <- read.table("linc_RNA/nonTADbound-lincRNA5.bed")
nTb_lincRNA10 <- read.table("linc_RNA/nonTADbound-lincRNA10.bed")
nTb_lincRNA20 <- read.table("linc_RNA/nonTADbound-lincRNA20.bed")
#loading non-TAD-bound pcgenes sets
nTb_pc5 <- read.table("pc_genes/nonTADbound-pcgene5.bed")
nTb_pc10 <- read.table("pc_genes/nonTADbound-pcgene10.bed")
nTb_pc20 <- read.table("pc_genes/nonTADbound-pcgene20.bed")

#adding colnames
colnames(Tb_lincRNA5)= colnames(Tb_lincRNA10)=colnames(Tb_lincRNA20)=colnames(nTb_lincRNA5)=colnames(nTb_lincRNA10)=
  colnames(nTb_lincRNA20)=colnames(Tb_pc5)=colnames(Tb_pc10)=colnames(Tb_pc20)=colnames(nTb_pc5)=colnames(nTb_pc10)=
  colnames(nTb_pc20) <-c("chr", "start", "end", "gene", "strand")
colnames(exp_pcgene)=colnames(exp_lincRNA) <- c("gene", "expression")

#Splitting expression levels data into TADbound (Tb) and non-TADbound (nTb)
exp_Tb_lincRNA5 <-exp_lincRNA[exp_lincRNA$gene %in% Tb_lincRNA5$gene,]
exp_Tb_lincRNA10 <-exp_lincRNA[exp_lincRNA$gene %in% Tb_lincRNA10$gene,]
exp_Tb_lincRNA20 <-exp_lincRNA[exp_lincRNA$gene %in% Tb_lincRNA20$gene,]

exp_nTb_lincRNA5 <-exp_lincRNA[exp_lincRNA$gene %in% nTb_lincRNA5$gene,]
exp_nTb_lincRNA10 <-exp_lincRNA[exp_lincRNA$gene %in% nTb_lincRNA10$gene,]
exp_nTb_lincRNA20 <-exp_lincRNA[exp_lincRNA$gene %in% nTb_lincRNA20$gene,]

exp_Tb_pc5 <-exp_pcgene[exp_pcgene$gene %in% Tb_pc5$gene,]
exp_Tb_pc10 <-exp_pcgene[exp_pcgene$gene %in% Tb_pc10$gene,]
exp_Tb_pc20 <-exp_pcgene[exp_pcgene$gene %in% Tb_pc20$gene,]

exp_nTb_pc5 <-exp_pcgene[exp_pcgene$gene %in% nTb_pc5$gene,]
exp_nTb_pc10 <-exp_pcgene[exp_pcgene$gene %in% nTb_pc10$gene,]
exp_nTb_pc20 <-exp_pcgene[exp_pcgene$gene %in% nTb_pc20$gene,]

#Writing into bed files for further use.
write.table(exp_Tb_lincRNA5,file = "exp_Tb_lincRNA5.txt",sep="\t",quote = F,col.names = F,row.names = F)
write.table(exp_Tb_lincRNA10,file = "exp_Tb_lincRNA10.txt",sep="\t",quote = F,col.names = F,row.names = F)
write.table(exp_Tb_lincRNA20,file = "exp_Tb_lincRNA20.txt",sep="\t",quote = F,col.names = F,row.names = F)

write.table(exp_nTb_lincRNA5,file = "exp_nTb_lincRNA5.txt",sep="\t",quote = F,col.names = F,row.names = F)
write.table(exp_nTb_lincRNA10,file = "exp_nTb_lincRNA10.txt",sep="\t",quote = F,col.names = F,row.names = F)
write.table(exp_nTb_lincRNA20,file = "exp_nTb_lincRNA20.txt",sep="\t",quote = F,col.names = F,row.names = F)

write.table(exp_Tb_pc5,file = "exp_Tb_pc5.txt",sep="\t",quote = F,col.names = F,row.names = F)
write.table(exp_Tb_pc10,file = "exp_Tb_pc10.txt",sep="\t",quote = F,col.names = F,row.names = F)
write.table(exp_Tb_pc20,file = "exp_Tb_pc20.txt",sep="\t",quote = F,col.names = F,row.names = F)

write.table(exp_nTb_pc5,file = "exp_nTb_pc5.txt",sep="\t",quote = F,col.names = F,row.names = F)
write.table(exp_nTb_pc10,file = "exp_nTb_pc10.txt",sep="\t",quote = F,col.names = F,row.names = F)
write.table(exp_nTb_pc20,file = "exp_nTb_pc20.txt",sep="\t",quote = F,col.names = F,row.names = F)

CData <-function(df){
  return(cbind(df,fact=rep(deparse(substitute(df)),length(df[,1]))))
}

whole_exp <- rbind(CData(exp_Tb_pc5), CData(exp_Tb_pc10), CData(exp_Tb_pc20), CData(exp_nTb_pc5), CData(exp_nTb_pc10), CData(exp_nTb_pc20),
                   CData(exp_Tb_lincRNA5), CData(exp_Tb_lincRNA10), CData(exp_Tb_lincRNA20), CData(exp_nTb_lincRNA5), CData(exp_nTb_lincRNA10), CData(exp_nTb_lincRNA20))

library(ggplot2)
library(gridExtra)
```
#TAD boundaries definition
When defining TAD boundaries as the areas 20% before and after TAD borders, those boundaries seemed to overlap with those of the neighbouring domains. To solve this issue, the TAD boundaries have been set to 3 different, arbitrary threshold: 5%, 10% and 20% of the TAD length.

#Overlapping genes with TAD boundaries
Both lincRNAs and protein coding genes have been overlapped with TAD boundaries, using the intersect program with -f 0.25 (genes must have at least 25% of their sequence overlapping a boundary to be considered). The set of overlaps was then cleaned by removing duplicates.

#Data

##lincRNA

Original set|Threshold|Overlaps|TAD-bound genes
------------|---------|--------|-----
2510|5%|865|609
||10%|1481|960
||20%|2554|1420

#Protein-coding genes

Original set|Threshold|Overlaps|TAD-bound genes
------------|---------|--------|-----
14846|5%|6576|4607
||10%|12017|7495
||20%|21544|11337

#Results

##Expression levels

```{r exp_lincRNA, echo=FALSE,fig.height=8,fig.width=6}
short_med<-function(x){return(round(median(log10(x)),3))}
short_wilcox <- function(x,y){return(round(wilcox.test(x, y)$p.value,5))}

l5 <-ggplot(data=whole_exp[whole_exp$fact %in% c("exp_Tb_lincRNA5","exp_nTb_lincRNA5"),])+
  geom_boxplot(aes(fact, log10(expression)), fill=c("darkred","darkblue"), notch = T)+theme_bw()+ggtitle("lincRNA, 5%")+
  annotate(x=c(1, 2),y=c(-1,-1),geom = "text", label=c(short_med(whole_exp$expression[whole_exp$fact=="exp_Tb_lincRNA5"]), 
                                                       short_med(whole_exp$expression[whole_exp$fact=="exp_nTb_lincRNA5"])))+
  scale_x_discrete(name=paste("p-value = ", short_wilcox(whole_exp$expression[whole_exp$fact=="exp_Tb_lincRNA5"],
                                                         whole_exp$expression[whole_exp$fact=="exp_nTb_lincRNA5"]), sep=" "),
                   labels=c("TADb", "non-TADb"))
l10 <-ggplot(data=whole_exp[whole_exp$fact %in% c("exp_Tb_lincRNA10","exp_nTb_lincRNA10"),])+
  geom_boxplot(aes(fact, log10(expression)), fill=c("darkred","darkblue"), notch = T)+theme_bw()+ggtitle("lincRNA, 10%")+
  annotate(x=c(1, 2),y=c(-1,-1),geom = "text", label=c(short_med(whole_exp$expression[whole_exp$fact=="exp_Tb_lincRNA10"]), 
                                                       short_med(whole_exp$expression[whole_exp$fact=="exp_nTb_lincRNA10"])))+
  scale_x_discrete(name=paste("p-value = ", short_wilcox(whole_exp$expression[whole_exp$fact=="exp_Tb_lincRNA10"],
                                                         whole_exp$expression[whole_exp$fact=="exp_nTb_lincRNA10"]), sep=" "),
                   labels=c("TADb", "non-TADb"))
l20 <-ggplot(data=whole_exp[whole_exp$fact %in% c("exp_Tb_lincRNA20","exp_nTb_lincRNA20"),])+
  geom_boxplot(aes(fact, log10(expression)), fill=c("darkred","darkblue"), notch = T)+theme_bw()+ggtitle("lincRNA, 20%")+
  annotate(x=c(1, 2),y=c(-1,-1),geom = "text", label=c(short_med(whole_exp$expression[whole_exp$fact=="exp_Tb_lincRNA20"]), 
                                                       short_med(whole_exp$expression[whole_exp$fact=="exp_nTb_lincRNA20"])))+
  scale_x_discrete(name=paste("p-value = ", short_wilcox(whole_exp$expression[whole_exp$fact=="exp_Tb_lincRNA20"],
                                                         whole_exp$expression[whole_exp$fact=="exp_nTb_lincRNA20"]), sep=" "),
                   labels=c("TADb", "non-TADb"))
p5 <-ggplot(data=whole_exp[whole_exp$fact %in% c("exp_Tb_pc5","exp_nTb_pc5"),])+
  geom_boxplot(aes(fact, log10(expression)), fill=c("darkred","darkblue"), notch = T)+theme_bw()+ggtitle("protein-coding, 5%")+
  annotate(x=c(1, 2),y=c(-1,-1),geom = "text", label=c(short_med(whole_exp$expression[whole_exp$fact=="exp_Tb_pc5"]), 
                                                       short_med(whole_exp$expression[whole_exp$fact=="exp_nTb_pc5"])))+
  scale_x_discrete(name=paste("p-value = ", short_wilcox(whole_exp$expression[whole_exp$fact=="exp_Tb_pc5"],
                                                         whole_exp$expression[whole_exp$fact=="exp_nTb_pc5"]), sep=" "),
                   labels=c("TADb", "non-TADb"))
p10 <-ggplot(data=whole_exp[whole_exp$fact %in% c("exp_Tb_pc10","exp_nTb_pc10"),])+
  geom_boxplot(aes(fact, log10(expression)), fill=c("darkred","darkblue"), notch = T)+theme_bw()+ggtitle("protein-coding, 10%")+
  annotate(x=c(1, 2),y=c(-1,-1),geom = "text", label=c(short_med(whole_exp$expression[whole_exp$fact=="exp_Tb_pc10"]), 
                                                       short_med(whole_exp$expression[whole_exp$fact=="exp_nTb_pc10"])))+
  scale_x_discrete(name=paste("p-value = ", short_wilcox(whole_exp$expression[whole_exp$fact=="exp_Tb_pc10"],
                                                         whole_exp$expression[whole_exp$fact=="exp_nTb_pc10"]), sep=" "),
                   labels=c("TADb", "non-TADb"))
p20 <-ggplot(data=whole_exp[whole_exp$fact %in% c("exp_Tb_pc20","exp_nTb_pc20"),])+
  geom_boxplot(aes(fact, log10(expression)), fill=c("darkred","darkblue"), notch = T)+theme_bw()+ggtitle("protein-coding, 20%")+
  annotate(x=c(1, 2),y=c(-1,-1),geom = "text", label=c(short_med(whole_exp$expression[whole_exp$fact=="exp_Tb_pc20"]), 
                                                       short_med(whole_exp$expression[whole_exp$fact=="exp_nTb_pc20"])))+
  scale_x_discrete(name=paste("p-value = ", short_wilcox(whole_exp$expression[whole_exp$fact=="exp_Tb_pc20"],
                                                         whole_exp$expression[whole_exp$fact=="exp_nTb_pc20"]), sep=" "),
                   labels=c("TADb", "non-TADb"))
grid.arrange(nrow=2, l5, l10, l20, p5, p10, p20)
```

##Subcellular localization

...

##Tissue specificity

...




